Modern Statistical Methods --- Example Sheet 1

### 2020-10-28 modern_statistical_methods/C2.tex
I. Hate. Matlab.

We have the following:
\begin{theorem}[Slow rate]
	Assume $X$ has centred and scaled columns, and assume that $Y$ has been centred, so $Y = X\beta^0 + \eps - \bar\eps \vec 1$. Let $A > 0$ and suppose
	\[
	\lambda = A\sigma\sqrt{\frac{\log(p)}{n}}.
	\]
	Let $\hat\beta = \hat\beta_\lambda^\Rm{L}$, then with probability at least $1 - 2p^{-A^2/2 - 1}$ we have that the MSPE satisfies
	\[
	\frac1n \norm{X(\beta^0 - \hat\beta)}_2^2 \leq 4\lambda \norm{\beta^0}_1 = 4A\sigma\sqrt{\frac{\log(p)}{n}} \norm{\beta^0}_1. 
	\]
\end{theorem}

\begin{proof}
	By definition we have
	\[
	\frac1{2n}\norm{Y - X\hat\beta }_2^2 + \lambda \norm{\hat \beta}_1 \leq \frac1{2n}\norm{Y - X\beta^0}_2^2 + \lambda \norm{\beta^0}_1,
	\]
	and rearranging the terms gives
	\[
	\frac1{2n} \norm{X(\beta^0 - \hat\beta)}_2^2 \leq\frac1n \eps\T X (\hat\beta - \beta^0) + \lambda\norm{\beta^0}_1 - \lambda \norm{\hat\beta}_1. 
	\]
	By H\"older's inequality we have $\abs{\eps\T X(\beta^0 - \hat\beta)} \leq \norm{X\T\eps}_\infty \norm{\hat\beta - \beta^0}_1$. 
	Define the event \\ $\Omega = \qty{\norm{X\T\eps}_\infty/n \leq \lambda}$, then conditional on $\Omega$ we find
	\[
	\frac1{n} \norm{X(\beta^0 - \hat\beta)}_2^2 \leq 2\lambda \qty(\norm{\beta^0 - \hat\beta}_1 + \norm{\beta^0}_1 - \norm{\hat\beta}_1) \leq 4\lambda \norm{\beta^0}_1.
	\]
	In (\TODO refer), we will show that $\PP(\Omega) \geq 1 - 2p^{-(A^2/2 - 1)}$, which completes the proof. 
\end{proof}

>>>>

We have the following:
\begin{theorem}[Slow rate]
	Assume $X$ has centred and scaled columns, and assume that $Y$ has been centred, so $Y = X\beta^0 + \eps - \bar\eps \vec 1$. Let $A > 0$ and suppose
	\[
	\lambda = A\sigma\sqrt{\frac{\log(p)}{n}}.
	\]
	Let $\hat\beta = \hat\beta_\lambda^\Rm{L}$, then with probability at least $1 - 2p^{-(A^2/2 - 1)}$ we have that the MSPE satisfies
	\[
	\frac1n \norm{X(\beta^0 - \hat\beta)}_2^2 \leq 4\lambda \norm{\beta^0}_1 = 4A\sigma\sqrt{\frac{\log(p)}{n}} \norm{\beta^0}_1. 
	\]
\end{theorem}

\begin{proof}
	By definition we have
	\[
	\frac1{2n}\norm{Y - X\hat\beta }_2^2 + \lambda \norm{\hat \beta}_1 \leq \frac1{2n}\norm{Y - X\beta^0}_2^2 + \lambda \norm{\beta^0}_1,
	\]
	and rearranging the terms gives
	\[
	\frac1{2n} \norm{X(\beta^0 - \hat\beta)}_2^2 \leq\frac1n \eps\T X (\hat\beta - \beta^0) + \lambda\norm{\beta^0}_1 - \lambda \norm{\hat\beta}_1. 
	\]
	By H\"older's inequality we have $\abs{\eps\T X(\beta^0 - \hat\beta)} \leq \norm{X\T\eps}_\infty \norm{\hat\beta - \beta^0}_1$. 
	Define the event \\ $\Omega = \qty{\norm{X\T\eps}_\infty/n \leq \lambda}$, then conditional on $\Omega$ we find
	\[
	\frac1{n} \norm{X(\beta^0 - \hat\beta)}_2^2 \leq 2\lambda \qty(\norm{\beta^0 - \hat\beta}_1 + \norm{\beta^0}_1 - \norm{\hat\beta}_1) \leq 4\lambda \norm{\beta^0}_1.
	\]
	In \cref{lem:probability_omega_bound}, we will show that $\PP(\Omega) \geq 1 - 2p^{-(A^2/2 - 1)}$, which completes the proof. 
\end{proof}

\subsection{Concentration inequalities}
Let $W$ be any random variable and $\phi \colon \RR\to[0,\infty)$ strictly increasing. Then by Markov's inequality we have
\[
\PP(W \geq t) = \PP(\phi(W) \geq \phi(t)) \leq \frac{\EE[\phi(W)]}{\phi(t)}. 
\]
Plugging in $\phi(x) = e^{\alpha x}$ (for some $\alpha > 0$), we get
\[
\PP(W \geq t) \leq e^{-\alpha t} \EE[e^{\alpha W}] = e^{-\alpha t} M_W(\alpha).
\]
Now we can take the infimum over all $\alpha$ on the right-hand side, and we get what is called the \emph{Chernoff bound}:
\[
\PP(W \geq t) \leq \inf_{\alpha > 0} e^{-\alpha t} M_W(\alpha). 
\]

\begin{definition}
	A random variable  $W$ with mean $\mu$ is called \emph{sub-Gaussian with parameter $\sigma > 0$} or $\sigma$-sub-Gaussian if 
	\[
	M_{W - \mu} \leq M_{N(0, \sigma^2)} \quad\text{or equivalently}\quad \EE[e^{\alpha(W - \mu)}] \leq e^{\alpha^2\sigma^2/2} \text{ for all $\alpha \in \RR$}. 
	\]
\end{definition}

We need the following lemma, which characterises an important class of sub-Gaussian random variables:
\begin{lemma}[Hoeffding]
	If $W$ is a mean-zero random variable which takes values in $[a, b]$, then $W$ is sub-Gaussian with parameter $(b-a)/2$. 
\end{lemma}

By Chernoff bounding, we obtain for a $\sigma$-sub-Gaussian random variable $W$ that \[\PP(W - \mu \geq t) \leq e^{-t^2/(2\sigma^2)}.\]

\begin{proposition}
	Let $W_1, \dotsc, W_n$ be independent, mean-zero random variables where $W_i$ is $\sigma_i$-sub-Gaussian. For any $\gamma \in \RR^n$, the random variable $\gamma\T W$ is sub-Gaussian with parameter $\qty(\sum_i \sigma_i^2\gamma_i^2)^{1/2}$.
\end{proposition}
\begin{proof}
	Since the $W_i$ are independent we have for all $\alpha \in \RR$ that
	\begin{align*}
		\EE[e^{\alpha \sum_i \gamma_i W_i}] = \prod_{i=1}^n \EE[e^{\alpha \gamma_i W_i}]  \leq \prod_{i=1}^n e^{\alpha^2\gamma_i^2\sigma_i^2/2} = e^{\alpha^2 \sum_i \gamma_i^2\sigma_i^2/2}.
	\end{align*}
\end{proof}

\begin{lemma} \label{lem:probability_omega_bound}
	Suppose $\eps_1, \dotsc, \eps_n$ are independent mean-zero $\sigma$-sub-Gaussian random variables and let $\lambda \ceq A\sigma \sqrt{\log(p)/n}$. Then
	\[
	\PP\qty(\frac{\norm{X\T\eps}_\infty}{n} \leq \lambda) \geq 1 - 2p^{-(A^2/2 - 1)}. 
	\]
\end{lemma}

\begin{proof}
	We have
	\[
	\PP\qty(\frac{\norm{X\T\eps}_\infty}{n} > \lambda) = \PP\qty(\bigcup_i \frac{\abs{X_i\T\eps}}{n} > \lambda) \leq \sum_i \PP\qty(\frac{\abs{X_i\T\eps}}{n} > \lambda). 
	\]
	By the previous proposition, both $X_i\T\eps$ and $-X_i\T\eps$ are sub-Gaussian with parameter $\sigma/\sqrt n$, so we have
	\[
	\sum_i \PP\qty(\frac{\abs{X_i\T\eps}}{n} > \lambda) \leq 2p \exp(\frac{-\lambda^2}{2(\sigma/\sqrt n)^2}) = 2p\exp(-A^2\log(p)/2) = 2 p^{-(A^2/2 - 1)}.
	\]
\end{proof}
